---
title: "Analysis of Genetic Data 1: Inferring Population Structure"
subtitle: (A Subtitle Would Go Here if This Were a Class)
author: Peter Carbonetto
institute: "Research Computing Center and 
   the Dept. of Human Genetics,
   University of Chicago"
titlegraphic: uchicago.png
fontsize: 12pt
output:
 beamer_presentation:
   template: beamer.tex
   keep_tex: false
 ioslides_presentation:
   smaller: false
make149: false
---

# Pop Songs and Political Science

## Sheena Easton and Game Theory

Sheena Easton describes the following scenario for her baby:

1. Takes the morning train
2. Works from nine 'til five
3. Takes another train home again
4. Finds Sheena Easton waiting for him

Sheena Easton and her baby are playing a **zero-sum (total conflict) game**.

- Akin to Holmes-Moriarty game (see: von Neumann and Morgenstern)
- Solution: **mixed strategy**

## Rick Astley's Re-election Platform

Rick Astley's campaign promises:

- Never gonna give you up.
- Never gonna let you down.
- Never gonna run around and desert you.
- Never gonna make you cry.
- Never gonna say goodbye.
- Never gonna tell a lie and hurt you.

Whereas these pledges conform to the preferences of the **median voter**, we expect Congressman Astley to secure re-election.

## Caribbean Queen and Operation Urgent Fury

Billy Ocean released "Caribbean Queen" in 1984.

- Emphasized sharing the same dream
- Hearts beating as one

"Caribbean Queen" is about the poor execution of Operation Urgent Fury.

- Echoed JCS chairman David Jones' frustrations with military establishment.

Billy Ocean is advocating for what became the Goldwater-Nichols Act.

- Wanted to take advantage of **economies of scale**, resolve **coordination problems** in U.S. military.

## The Good Day Hypothesis

We know the following about Ice Cube's day.

1. The Lakers beat the Supersonics.
2. No helicopter looked for a murder.
3. Consumed Fatburger at 2 a.m.
4. Goodyear blimp: "Ice Cube's a pimp."

This leads to two different hypotheses:

- $H_0$: Ice Cube's day is statistically indistinguishable from a typical day.
- $H_1$: Ice Cube is having a good (i.e. greater than average) day.

These hypotheses are tested using archival data of Ice Cube's life.

# Example R code
## Example R stuff

```{r cars, echo = TRUE}
summary(cars)
```

## Slide with Plot

```{r pressure}
plot(pressure)
```

## ggplot code

```{r createdata, echo=TRUE, eval=FALSE}
df <- data.frame(x = rnorm(1000))
x <- df$x
base <- ggplot(df, aes(x)) + geom_density()  + scale_x_continuous(limits = c(-5, 5))
base + stat_function(fun = dnorm, colour = "red")
```


## Another Plot

```{r plotit, echo=FALSE}
library(ggplot2)
df <- data.frame(x = rnorm(1000))
x <- df$x
base <- ggplot(df, aes(x)) + geom_density()  + scale_x_continuous(limits = c(-5, 5))
base + stat_function(fun = dnorm, colour = "red")
```